---
title: "rgts"
output: rmarkdown::html_vignette
description:
  An introduction to accessing GridTimeSeries (GTS) data with rgts.
vignette: >
  %\VignetteIndexEntry{rgts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

rgts is an R package which makes it easier to download the GridTimeSeries (GTS) data available in a 1x1 km grid from NVE. The package use well known R syntax to connect to the data base, downloads data and casts it into easily accessible data.frames. Each row contains the data of a grid cell at a given date and time.

Below is an overview of how to use the package. This brief introduction assumes that you are familiar with working with spatial data in R. We make use of the [sf](https://r-spatial.github.io/sf/index.html) and [stars](https://r-spatial.github.io/stars/index.html) packages. Consult these packages as needed. First we load the `rgts`, `sf` and `stars` packages.

```{r setup}
library(rgts)
library(sf)
library(stars)
```

## Types of input
To download GTS data you will need one out of three possible types of spatial data as input to the functions.

* Points with coordinates
* An area represented as a polygon
* Cellindices

The first two supposedly do not require too much explanation, either you download data for one or more points of interest (e.g. the position of an animal) or for one or more areas of interest (e.g. the outline of a protected area). The point and polygon data should be `sf` objects.

The third on the other hand is less clear. In the GTS database, each 1x1 km grid cell in Norway is assigned a unique ID, called cellindex. This index starts on 0 for the cell in the north-western corner and numbers cells continuously left to right one row at the time. The function `gts_create_grid()` builds a stars raster with the same extent and resolution as the GTS data and cellindices as an attribute. Below we create the grid and extract the cellindex for cell numbers 1 to 5 in the `stars` raster.

```{r grid}
## Build raster
grid <- gts_create_grid()
grid
## Extract the cellindex of the first five cells
grid$cellindex[1:5]
```

What you see is that the cellindex starts at 0 and is equal to (cell number - 1) in the raster. This is important to remember if we later want to plot values on a map.

## Working with cellindices
How do we obtain cellindices then? If you have a point or area of interest, the function `gts_cellindex()` is used to extract the cellindices for the grid cells which contain a point or which are within an area. What you need as input to the function is the geometry, point(s) or polygon(s), as `sf` objects. The approach for polygons and points are the same. For points this can be done like like this:

```{r cellindex}
## Create a data frame with coordinates
coords <- data.frame(x = c(58200, 74225), y = c(6708225, 6715025))
coords <- st_as_sf(x = coords, coords = c("x", "y"), crs = 25833)
## Extract cellindices
gts_cellindex(geometry = coords)
```

These cellindices can later be used to query for GTS data. Generally, performing queries using cellindices as input is faster than using coordinates or polygons. This is because the latter two requires the API to search for cellindices before looking up data.

## Coordinate system
The GTS data grid is in the coordinate system (CRS) ETRS89 UTM 33N ([EPSG: 25833](https://epsg.io/25833)). Hence, any points or polygons used as input also needs to use the same CRS. CRS transformations are straight forward using the `sf` package and is not covered here.

## Finding parameters
There are several weather parameters available for download in the GTS data. These come in three different time resolutions, 1440, 180 and 60 minutes. Daily estimates (1440 minutes), 3 hour and 1 hour estimates have different temporal availability. Some of the daily parameters go all the back to 1957, while the 3 and 1 hour parameters are available from 2010 or more recent years. To find parameters and their temporal availability use the function `gts_parameters` to read in a table from the API.

```{r param}
## Find the first five parameters
head(gts_parameters(), 5)

```


## Download 1: Coordinates
For a set of points of interest, data for grid cells which contain these points can be downloaded.

```{r points}
## Create a data frame with coordinates
coords <- data.frame(x = c(58200, 74225), y = c(6708225, 6715025))
## Download data
gts_dl_coords(coords = coords, parameter = "tm", start_date = "2023-12-01", end_date = "2023-12-02")

```

The output is a data frame where the coordinates we supplied are returned, as well as the corresponding cellindex and the average altitude in the grid cell. The data for these grid cells are given for the specified dates, time and time resolution. Units provide the units of measurement and the column `tm` contains the data for the parameter of interest. The last column is always named by the abbreviated parameter name in the query.

## Download 2: Polygons
Lets create a polygon, which covers an area of interest at Hardangervidda in Norway.

```{r poly}
## Create a sf polygon
library(sf)
xx <- c(58200, 74225, 58200, 74226)
yy <- c(6708225, 6708225, 6715025, 6715025)
points <- st_as_sf(x = data.frame(x = xx, y = yy), coords = c("x", "y"), crs = 25833)
pol <- st_as_sf(st_as_sfc(st_bbox(points)))
```

We can now take this polygon and download daily temperature data ("tm") for all grid cells within the polygon. First we need to convert the polygon to json using the function `sf_to_json`. We only show the first 10 lines of data.

```{r download poly}
## Convert to json
pol_json <- gts_sf2json(pol)
## Download data
head(gts_dl_polygon(polygon = pol_json, parameter = "tm", start_date = "2023-12-01", end_date = "2023-12-01"), 10)
```

## Download 3: Cellindices
It is faster to query the API using cellindices directly instead of going the way around with polygons. First, the cellindices for a given polygon is obtained by

```{r cellindex poly}
cells <- gts_cellindex(geometry = pol)
```

Then the daily temperature data can be downloaded using the `gts_dl_cellindex`-function. Again we only show the first 10 lines of data.

```{r download cellindex}
## Download data
head(gts_dl_cellindex(cellindex = cells, parameter = "tm", start_date = "2023-12-01", end_date = "2023-12-01"), 10)
```

We can also obtain data aggregated across grid cells. We then need to choose a method for aggregating data, "sum", "min", "max", "avg" or "median". Choosing "avg" we can obtain daily averages by

```{r download cellindex aggr}
## Download data
gts_dl_cellindex_aggr(cellindex = cells, parameter = "tm", start_date = "2023-12-01", end_date = "2023-12-10", method = "avg")
```

It is also possible to use `gts_dl_polygon_aggr`-function to downloading aggregated data using polygons directly, but remember to convert the polygon to json first. As mentioned, the polygon functions are a bit slower, so use the cellindex-functions when speed is important.

## Maps
Work in progress...

## Troubleshooting
Sometimes what seems to be a valid query may fail and we need to do some troubleshooting. The download functions, with prefix `gts_dl_...`, have two optional arguments to aid this process. First, you may set `verbose` to `TRUE` to see exactly what is sent to the server. Then you can check that you have provided the input data correctly. Second, you can set `return_raw` to `TRUE` to have the function return the raw result as a `list` from the query. Then check that the output is as expected from the type of data and time span and resolution in the query.

Here is how to handle a few specific recurring issues when using the `gts_dl_...`-functions which results in error messages from the server:

* Multiple coordinates for `points` in the same grid cell.
  + The API do no accept queries which ask for the data from the same grid cell multiple times. Find the cellindices using `gts_cellindex` and perform the query with unique cellindices using `gts_dl_cellindex` or `gts_dl_cellindex_aggr`.
* Too large queries.
  + The API have a upper limit for the size of data which can be returned in each query. Generally the cap is at around 635'000 values, but this number decreases with increasing number of missing values. Run the query in chunks to limit the number of cellindices or the length of the time series in each query to get around this issue.
* Missing data in database.
  + Sometimes the API returns an error message if the query requests data outside the available time spans for data or there is missing data in the database.
  + Check the available time spans for data using `gts_parameters` to make sure that the query is valid.
  + Next, try some shorter time spans to figure out if there are particular dates which have missing data.
  + Change to the `gts_dl_cellindex` and `gts_dl_polygon` functions, which are less likely to return error messages for missing data than the `gts_dl_cellindex_aggr` and `gts_dl_polygon_aggr` functions. This is because of the way that the API handles missing data. For the former two functions the API return a NoData value, while for the latter two the API currently ends with an error.

## Getting help
If you encounter a clear bug, please file a minimal reproducible example on [github](https://github.com/thomas-kvalnes/rgts/issues). Questions about the data needs to be addressed to [NVE](https://www.nve.no/om-nve/kontakt/). General R help can be found in the R Community on [Stack Overflow](https://stackoverflow.com/collectives/r-language). E-mail the maintainer for urgent or private communications.

